KiCS2: A New Compiler from Curry to Haskell

### 2011-05-09 papers/wflp2011/paper.tex
Paper fine-tuning

\begin{abstract}
In this paper we present our first steps towards a new system
to compile functional logic programs of the source language Curry
into purely functional Haskell programs.
Our implementation is based on the idea to represent
non-deterministic results as values in the data types corresponding
to the results.
This enables the application of various search strategies
to extract values from the search space.
We show by several benchmarks that our implementation
can compete with or outperforms other existing implementations of Curry.
\end{abstract}

@@

The implementation of functional logic languages is challenging
since a reasonable implementation has to support the operational
features mentioned above.
One possible approach
is the design of new abstract machines appropriately supporting
these operational features and implementing them in some (typically,
imperative) language, like C \cite{Lux99FLOPS}
or Java \cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP}.
Another approach is the reuse of already existing implementations
of some of these features by translating
functional logic programs into logic or functional languages.
For instance, if one compiles into Prolog, one can reuse
the existing backtracking implementation for non-deterministic
search as well as logic variables and unification for computing with partial
information. However, one has to implement demand-driven evaluation
and higher-order functions \cite{AntoyHanus00FROCOS}.
A disadvantage of this approach is the commitment to a fixed
search strategy (backtracking).

@@

The general objective of our approach is the support
of flexible strategies to explore the search space
resulting from non-deterministic computations.
In contrast to Prolog-based implementations
that use backtracking and, therefore, are incomplete,
we also want to support complete strategies like breadth-first search,
iterative deepening or parallel search (in order to exploit
multi-core architectures). We achieve this goal by an
explicit representation of the search space as data
that can be traversed by various operations.
Moreover, purely deterministic computations
are implemented as purely functional programs so that
they are executed with almost the same efficiency
of their purely functional counterparts.

@@

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{haskell}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                 $\to~$  xor True False      $\to~$  True
\end{haskell}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated before the function calls.
In order to avoid dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and non-deterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics}
where values of the arguments of an operation are determined before the
operation is evaluated. Note that this does not necessarily require
an eager evaluation of arguments.
Actually, \cite{AlbertHanusHuchOliverVidal05,LopezRodriguezSanchez07}
define lazy evaluation strategies for functional logic programs
with call-time choice semantics where actual arguments passed to
operations are shared. Hence, we can evaluate the expression above
lazily, provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}.
The requirements of this call-time choice semantics are the
reason why it is not simply possible to use list comprehensions
or non-determinism monads for a straightforward implementation
of functional logic programs in Haskell \cite{FischerKiselyovShan09}.

@@

\begin{figure}[t]
\[
\begin{array}{@{}rcl@{~~~~~~~~}l}
  e & ::= & x & \mbox{$x$ is a variable}\\
    &  |  & c(e_1,\ldots,e_n) & \mbox{$c$ is an $n$-ary constructor symbol}\\
    &  |  & f(e_1,\ldots,e_n) & \mbox{$f$ is an $n$-ary function symbol}\\
    &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\[1ex]
  D & ::= & f(x_1,\ldots,x_n) = e
                & \mbox{$n$-ary function $f$ with a single rule}\\
    &  |  & f(c(y_1,\ldots,y_m),x_2,\ldots,x_n) = e
                & \mbox{matching rule for $n$-ary function $f$}\\
    &     & & \mbox{$c$ is an $m$-ary constructor symbol}\\
  P & ::= & D_1 \ldots D_k
\end{array}
\]
\caption{Uniform programs ($e$: expressions, $D$: definitions, $P$: programs)}
\label{fig:uniform}
\end{figure}
%
Finally, we assume that the pattern matching strategy
is explicitly encoded in individual matching functions.
In contrast to \cite{AlbertHanusHuchOliverVidal05},
where the pattern matching strategy is encoded in case expressions,
we assume that each case expression is transformed into
a new operation in order to avoid complications by translating
nested case expressions.
Thus, we assume that all programs are \emph{uniform}
according to the definition in Fig.~\ref{fig:uniform}.\footnote{%
A variant of uniform programs has been considered in
\cite{MorenoEtAl90ALP} to define lazy narrowing strategies
for functional logic programs. Although the motivations are similar,
our notion of uniform programs is more restrictive since we
allow only a single non-variable argument in each left-hand side
of a rule.
Uniform programs have also been applied in
\cite{Zartmann97SAS} to define a denotational analysis
of functional logic programs.}
There, the variables in the left-hand sides of each rule are pairwise
different, and
the constructors in the left-hand sides of the matching rules of each function
are pairwise different.
Uniform programs have a simple form of pattern matching:
either a function is defined by a single rule without pattern matching,
or it is defined by rules with only one constructor in the left-hand
side of each rule, and in the same argument for all rules.%
\footnote{For simplicity, we require in
Fig.~\ref{fig:uniform} that the matching argument is always the
first one, but one can also choose any other argument.}
For instance, the operation \code{xor} defined above can be transformed
into the following uniform program:
%
\begin{curry}
  xor True   x = xor' x
  xor False  x = x
  xor' False = True
  xor' True  = False
\end{curry}
%
In particular, there are no overlapping rules for functions
(except for the choice operation \ccode{?} which is considered as predefined).
Antoy \cite{Antoy01PPDP} showed that each functional logic program,
i.e., each constructor-based conditional term rewriting system,
can be translated into an equivalent unconditional term rewriting system
without overlapping rules but containing choices in the right-hand sides,
also called LOIS (limited overlapping inductively sequential) system.
Furthermore, Bra{\ss}el \cite{Brassel11Thesis} showed the semantical
equivalence of narrowing computations in LOIS systems
and rewriting computations in uniform programs.
Due to these results, uniform programs are a reasonable intermediate
language for our translation into Haskell which will be presented
in the following.

@@

We can correct this unintended behavior of our implementation
by identifying different \code{Choice} occurrences that are duplicates
of some single \code{Choice}. This can be easily done by attaching
a unique identifier, e.g., a number, to each choice:
\begin{haskell}
  type ID = Integer
  data Bool = False | True | Choice ID Bool Bool
\end{haskell}
Furthermore, we modify the \code{Choice} pattern rules so that
the identifiers will be kept, e.g.,
\begin{haskell}
  xor (Choice i x1 x2) x = Choice i (xor x1 x) (xor x2 x)
\end{haskell}
If we evaluate the expression \ccode{xorSelf aBool} and
assign the number \code{1}
to the choice of \code{aBool}, we obtain the result
\begin{haskell}
  Choice 1 (Choice 1 False True) (Choice 1 True False)
\end{haskell}
When we show the values contained in this result,
we have to make \emph{consistent} selections in choices with
same identifiers. This, if we select the left branch as the value
of the outermost \code{Choice}, we also have to select the left branch
in the selected argument \code{(Choice 1 False True)} so that only
the value \code{False} is possible here.
Similarly, if we select the right branch as the value of the outermost
\code{Choice}, we also have to select the right branch in
its selected argument \code{(Choice 1 True False)} which yields the only
value \code{False}.

@@

Functional logic computations can also fail, e.g.,
due to partially defined operations.
Computing with failures is a typical programming technique
and provides for specific programming patterns \cite{AntoyHanus02FLOPS}.
Hence, in contrast to functional programming,
a failing computation should not abort the complete evaluation
but it should be considered as some part of a computation that
does not produce a meaningful result.
In order to implement this behavior, we extend
each datatype by a further constructor \code{Fail}
and complete each function containing matching rules
by a final rule that matches everything and returns the value \code{Fail}.
For instance, consider the definition of lists
\begin{curry}
  data List a = Nil | Cons a (List a)
\end{curry}
and an operation to extract the first element of a non-empty list:
\begin{curry}
  head :: List a -> a
  head (Cons x xs) = x
\end{curry}
The type definition is extended as follows:\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data List a = Nil | Cons a (List a) | Choice (List a) (List a) | Fail
\end{haskell}
The operation \code{head} is extended by an identifier supply
and further matching rules:
\begin{haskell}
  head :: List a -> IDSupply -> a
  head (Cons x xs)      s = x
  head (Choice i x1 x2) s = Choice i (head x1 s) (head x2 s)
  head _                s = Fail
\end{haskell}
Note that the final rule returns \code{Fail} if \code{head} is
applied to the empty list as well as if the matching argument
is already a failed computation, i.e., it also propagates failures.

@@

In this section we evaluate our compiler by comparing the
efficiency of the generated Haskell programs to various
other systems, in particular, other implementations of Curry.
For our comparison with other Curry implementations,
we consider PAKCS \cite{Hanus10PAKCS} (Version 1.9.2) which compiles
Curry into Prolog \cite{AntoyHanus00FROCOS}
(based on SICStus-Prolog 4.1.2, a SWI-Prolog 5.10 back end is also
available
but much slower). PAKCS has been used for a number of practical
applications of Curry.
Another mature implementation which we consider is MCC \cite{Lux99FLOPS}
(Version 0.9.10) which compiles Curry into C.
MonC \cite{BrasselFischerHanusReck11} is a compiler
from Curry into Haskell. It is based on a monadic representation
of non-deterministic computations where sharing is explicitly
managed by the technique proposed in \cite{FischerKiselyovShan09}.
Since this compiler
is in an experimental state, we could not execute all benchmarks with
MonC (these are marked by ``n/a'').

@@

Our compiler has been executed with the Glasgow Haskell Compiler
(GHC 6.12.1, option -O2). All benchmarks were executed on a Linux machine
running Debian 5.0.7 with an Intel Core 2 Duo (3.0GHz) processor.
The timings were performed with the time command measuring the
execution time (in seconds) of a compiled executable for each benchmark
as a mean of three runs. ``oom'' denotes a memory overflow in a computation.

>>>>

\begin{abstract}
In this paper we present our first steps towards a new system
to compile functional logic programs of the source language Curry
into purely functional Haskell programs.
Our implementation is based on the idea to represent
non-deterministic results as values of the data types corresponding
to the results.
This enables the application of various search strategies
to extract values from the search space.
We show by several benchmarks that our implementation
can compete with or outperform other existing implementations of Curry.
\end{abstract}

@@

The implementation of functional logic languages is challenging
since a reasonable implementation has to support the operational
features mentioned above.
One possible approach
is the design of new abstract machines appropriately supporting
these operational features and implementing them in some (typically,
imperative) language, like C \cite{Lux99FLOPS}
or Java \cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP}.
Another approach is the reuse of already existing implementations
of some of these features by translating
functional logic programs into either logic or functional languages.
For instance, if one compiles into Prolog, one can reuse
the existing backtracking implementation for non-deterministic
search as well as logic variables and unification for computing with partial
information. However, one has to implement demand-driven evaluation
and higher-order functions \cite{AntoyHanus00FROCOS}.
A disadvantage of this approach is the commitment to a fixed
search strategy (backtracking).

@@

The general objective of our approach is the support
of flexible strategies to explore the search space
resulting from non-deterministic computations.
In contrast to Prolog-based implementations
that use backtracking and, therefore, are incomplete,
we also want to support complete strategies like breadth-first search,
iterative deepening or parallel search (in order to exploit
multi-core architectures). We achieve this goal by an
explicit representation of the search space as data
that can be traversed by various operations.
Moreover, purely deterministic computations
are implemented as purely functional programs so that
they are executed with almost the same efficiency
as their purely functional counterparts.

@@

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{haskell}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                 $\to~$  xor True False      $\to~$  True
\end{haskell}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated before the operation calls.
In order to avoid dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and non-deterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics}
where values of the arguments of an operation are determined before the
operation is evaluated. Note that this does not necessarily require
an eager evaluation of arguments.
Actually, \cite{AlbertHanusHuchOliverVidal05,LopezRodriguezSanchez07}
define lazy evaluation strategies for functional logic programs
with call-time choice semantics where actual arguments passed to
operations are shared. Hence, we can evaluate the expression above
lazily, provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}.
The requirements of the call-time choice semantics are the
reason why it is not simply possible to use list comprehensions
or non-determinism monads for a straightforward implementation
of functional logic programs in Haskell \cite{FischerKiselyovShan09}.

@@

\begin{figure}[t]
\[
\begin{array}{@{}rcl@{~~~~~~~~}l}
  e & ::= & x & \mbox{$x$ is a variable}\\
    &  |  & c(e_1,\ldots,e_n) & \mbox{$c$ is an $n$-ary constructor symbol}\\
    &  |  & f(e_1,\ldots,e_n) & \mbox{$f$ is an $n$-ary function symbol}\\
    &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\[1ex]
  D & ::= & f(x_1,\ldots,x_n) = e
                & \mbox{$n$-ary function $f$ with a single rule}\\
    &  |  & f(c(y_1,\ldots,y_m),x_2,\ldots,x_n) = e
                & \mbox{matching rule for $n$-ary function $f$}\\
    &     & & \mbox{$c$ is an $m$-ary constructor symbol}\\
  P & ::= & D_1 \ldots D_k
\end{array}
\]
\caption{Uniform programs ($e$: expressions, $D$: definitions, $P$: programs)}
\label{fig:uniform}
\end{figure}
%
Finally, we assume that the pattern matching strategy
is explicitly encoded in individual matching functions.
In contrast to \cite{AlbertHanusHuchOliverVidal05},
where the pattern matching strategy is encoded in case expressions,
we assume that each case expression is transformed into
a new operation in order to avoid complications arising from the translation
of nested case expressions.
Thus, we assume that all programs are \emph{uniform}
according to the definition in Fig.~\ref{fig:uniform}.\footnote{%
A variant of uniform programs has been considered in
\cite{MorenoEtAl90ALP} to define lazy narrowing strategies
for functional logic programs. Although the motivations are similar,
our notion of uniform programs is more restrictive since we
allow only a single non-variable argument in each left-hand side
of a rule.
Uniform programs have also been applied in
\cite{Zartmann97SAS} to define a denotational analysis
of functional logic programs.}
There, the variables in the left-hand sides of each rule are pairwise
different, and
the constructors in the left-hand sides of the matching rules of each function
are pairwise different.
Uniform programs have a simple form of pattern matching:
either a function is defined by a single rule without pattern matching,
or it is defined by rules with only one constructor in the left-hand
side of each rule, and in the same argument for all rules.%
\footnote{For simplicity, we require in
Fig.~\ref{fig:uniform} that the matching argument is always the
first one, but one can also choose any other argument.}
For instance, the operation \code{xor} defined above can be transformed
into the following uniform program:
%
\begin{curry}
  xor True   x = xor' x
  xor False  x = x
  xor' False = True
  xor' True  = False
\end{curry}
%
In particular, there are no overlapping rules for functions
(except for the choice operation \ccode{?} which is considered as predefined).
Antoy \cite{Antoy01PPDP} showed that each functional logic program,
i.e., each constructor-based conditional term rewriting system,
can be translated into an equivalent unconditional term rewriting system
without overlapping rules but containing choices in the right-hand sides,
also called LOIS (limited overlapping inductively sequential) system.
Furthermore, Bra{\ss}el \cite{Brassel11Thesis} showed the semantical
equivalence of narrowing computations in LOIS systems
and rewriting computations in uniform programs.
Due to these results, uniform programs are a reasonable intermediate
language for our translation into Haskell which will be presented
in the following.

@@

We can correct this unintended behavior of our implementation
by identifying different \code{Choice} occurrences that are duplicates
of some single \code{Choice}. This can be easily done by attaching
a unique identifier, e.g., a number, to each choice:
\begin{haskell}
  type ID = Integer
  data Bool = False | True | Choice ID Bool Bool
\end{haskell}
Furthermore, we modify the \code{Choice} pattern rules so that
the identifiers will be kept, e.g.,
\begin{haskell}
  xor (Choice i x1 x2) x = Choice i (xor x1 x) (xor x2 x)
\end{haskell}
If we evaluate the expression \ccode{xorSelf aBool} and
assign the number \code{1}
to the choice of \code{aBool}, we obtain the result
\begin{haskell}
  Choice 1 (Choice 1 False True) (Choice 1 True False)
\end{haskell}
When we show the values contained in this result,
we have to make \emph{consistent} selections in choices with
same identifiers. Thus, if we select the left branch as the value
of the outermost \code{Choice}, we also have to select the left branch
in the selected argument \code{(Choice 1 False True)} so that only
the value \code{False} is possible here.
Similarly, if we select the right branch as the value of the outermost
\code{Choice}, we also have to select the right branch in
its selected argument \code{(Choice 1 True False)} which yields the sole
value \code{False}.

@@

Functional logic computations can also fail, e.g.,
due to partially defined operations.
Computing with failures is a typical programming technique
and provides for specific programming patterns \cite{AntoyHanus02FLOPS}.
Hence, in contrast to functional programming,
a failing computation should not abort the complete evaluation
but it should be considered as some part of a computation that
does not produce a meaningful result.
In order to implement this behavior, we extend
each data type by a further constructor \code{Fail}
and complete each operation containing matching rules
by a final rule that matches everything and returns the value \code{Fail}.
For instance, consider the definition of lists
\begin{curry}
  data List a = Nil | Cons a (List a)
\end{curry}
and an operation to extract the first element of a non-empty list:
\begin{curry}
  head :: List a -> a
  head (Cons x xs) = x
\end{curry}
The type definition is extended as follows:\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data List a = Nil | Cons a (List a) | Choice (List a) (List a) | Fail
\end{haskell}
The operation \code{head} is extended by an identifier supply
and further matching rules:
\begin{haskell}
  head :: List a -> IDSupply -> a
  head (Cons x xs)      s = x
  head (Choice i x1 x2) s = Choice i (head x1 s) (head x2 s)
  head _                s = Fail
\end{haskell}
Note that the final rule returns \code{Fail} if \code{head} is
applied to the empty list as well as if the matching argument
is already a failed computation, i.e., it also propagates failures.

@@

In this section we evaluate our compiler by comparing the
efficiency of the generated Haskell programs to various
other systems, in particular, other implementations of Curry.
For our comparison with other Curry implementations,
we consider PAKCS \cite{Hanus10PAKCS} (Version 1.9.2) which compiles
Curry into Prolog \cite{AntoyHanus00FROCOS}
(based on SICStus-Prolog 4.1.2, a SWI-Prolog 5.10 back end is also
available
but much slower). PAKCS has been used for a number of practical
applications of Curry.
Another mature implementation we consider is MCC \cite{Lux99FLOPS}
(Version 0.9.10) which compiles Curry into C.
MonC \cite{BrasselFischerHanusReck11} is a compiler
from Curry into Haskell. It is based on a monadic representation
of non-deterministic computations where sharing is explicitly
managed by the technique proposed in \cite{FischerKiselyovShan09}.
Since this compiler
is in an experimental state, we could not execute all benchmarks with
MonC (these are marked by ``n/a'').

@@

Our compiler has been executed with the Glasgow Haskell Compiler
(GHC 6.12.3, option -O2). All benchmarks were executed on a Linux machine
running Debian 5.0.7 with an Intel Core 2 Duo (3.0GHz) processor.
The timings were performed with the time command measuring the
execution time (in seconds) of a compiled executable for each benchmark
as a mean of three runs. ``oom'' denotes a memory overflow in a computation.
